Using numerical modelling to investigate temperature and wave exposure as drivers

### 2019-11-28 morphology_paper.Rmd
grammer and spelling recommendations added. Updated figures and extra paragraph to follow

---
title: Using numerical modelling to investigate temperature and wave exposure as drivers
  of macroalgal morphological characteristics
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
bibliography: references.bib
output:
  html_document:
    df_print: paged
  pdf_document:
    fig_caption: yes
  word_document:
    reference_docx: frontiers_template.docx
geometry: margin=1in
fontsize: 10pt
papersize: A4
tables: yes
abstract: 'Macroalgae morphological characteristics are determined by
  a combination of environmental factors with wave exposure and temperature being
  identified as the main drivers of morphological adaptation. Macroalgae have shown
  to exhibit different responses to different magnitudes of exposure, such as reduction
  in overall size and strength increasing traits. In terms of temperature, warmer
  environments have been shown to reduce the overall size of resident and transplanted
  species. However, none of the past studies have identified specific wave and temperature
  metrics responsible for the morphological adaptation macroalgae exhibit. Past research
  has often used simple or two dimensional models of wave exposure which does not
  take into account important aspects of the nearshore environment such as wave breaking,
  refraction and diffraction. Furthermore, past studies have often used satellite
  as sources for temperature data which has been shown to have large bias when measuring
  the nearshore environment. This study used *in situ* temperature data and wave power
  calculated from a 3D-numerical model to identify specific temperature and wave metrics
  responsible for morphological adaptation.  The results identify wave exposure as
  the main driver of morphological adaptation while identifying specific wave metrics.
  Furthermore, the results show differences in wave metrics between species; and between
  deep and shallow populations.'
---

@@

Kelps are a group of large seaweeds of the order Laminariales (Ochrophyta), which despite their relatively low taxonomic diversity of species in genera [@Bolton2010], form the basis of one of the most productive ecosystems globally [@Mann1973].  Kelps generally have a dependence on cool-temperate and arctic seawater temperatures [@Bolton2010; @Mohring2014-bk; @Cavanaugh2011-ib; @Dayton1985], and dominate the nearshore biomass within the rocky shallow coasts in both hemispheres [@Steneck2002]. Their size and complex morphology provide a heterogeneous habitat structure [@Steneck2002] that accommodate a multitude of turf and sub-canopy seaweed species, and diverse assemblages of sessile and mobile invertebrates and vertebrates [@Mann1973; @Steneck2002]; each depending on a wide suite of ecological services provided by kelp forests [@Gaines1987-bg; @Paul1993-bm; @Levin1994-qf; @Willis2003-xo; @anderson1997; @Dayton1985]. Wave exposure and temperature are regarded as important environmental drivers of kelp forests, and play a role in the distribution [@Gorman2013-nj], abundance [@Cavanaugh2011-ib; @Dayton1998-xw], diversity [@Wing2007-tv; @Wernberg2008], composition [@Leliaert2000a; @Dayton1985; @Norderhaug2012-ks; @Harley2012], growth [@Cousens1982-ck] and productivity [@Pedersen2012-vr; @Dayton1998-xw; @Krumhansl2012] of kelps.

Given kelps are cool-temperate organisms which are vulnerable to dislodgement under high wave exposure scenarios, changes in ocean temperature and an increase in storm frequency and magnitude pose the biggest direct threats to their survival. Warmer temperatures reduces the resilience of kelp individuals to storm disturbance [@wernberg2010], causes fragmentation through weakening tissues [@Simonson2015] and reduces growth and productivity [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]; while storms and high wave energy dislodge and break kelps [@byrnes2011; @graham2004; @seymour1989]. Despite these threats and disturbances kelps persist across a broad range of environments which is largely due to their morphological plasticity [@FowlerWalker2006; @wernberg2010]. A study by @Wernberg2003-mp investigated the morphology of *E. radiata* in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australasian coast. They found no correlation between spatial distance and morphological similarity, and rather the morphology of kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales [@Wernberg2003-mp]. Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Deep and shallow water environments may differ in abiotic processes such as the degree of water column mixing [@Smit2013a], solar heating [@Dellatorre2012; @Dunne2001] and effects of wave dampening [@Kobayashi1993-os; Mia2013; @Dubi1995].

Although multiple environmental factors play a role in kelp morphological adaptation, wave exposure and temperature have been identified as the main drivers across various kelp species [@bekkby2014; @Wernberg2005c; @Denny2002; @Hurd2000; @FowlerWalker2005; @Thomsen2004; @mabin2013]. For example, a study by @Serisawa2002-dc compared the morphology of *Ecklonia cava* Kjellman growing at warmer and cooler sites, and showed wrinkling of the blade and reduced size to be a characteristic of higher temperature ranges. Warmer temperature effects important kelp physiological processes such as photosynthesis and respiration, which influences growth and productivity and ultimately leads to a reduction in size [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]. Past research has shown that in highly wave exposed areas kelp morphology tends to take on characteristics which reduce drag, increase strength and increase flexibility [@Denny2002; @Hurd2000; @FowlerWalker2005; @Thomsen2004]. For example, a study by @Wernberg2005c examined the consistency of wave exposure as a driver of *Ecklonia radiata*  (C. Agardh) J. Agardh. across a broad geographic range, and showed trends towards drag-reducing (small size, narrow laterals and blades, low spinosity) and increased strength (large holdfast, thick stipe and thick blades and lamina) at high wave exposure sites.

Measures of wave exposure in ecological studies often incorporate integrative measures of hydrodynamic conditions at a particular site which are based on cartographical models of exposure. Cartographical models of wave exposure are 'fetch-based-models' which measure the length of open water associated with a particular site in a straight line, and are regarded as simple measures of wave exposure. Advances in numerical modelling based on physical/linear wave theory incorporate more complexity (wind forcing, wave-wave interactions, wave breaking, diffraction and variation in wave direction) into the models and allow for a quantitative, reproducible approach for characterising the hydrodynamic environment. Although @mabin2013 investigated kelp morphological characteristics using a 2D WAM model, currently no studies investigating macroalgal morphological characteristics in relation to the wave environment have incorporated 3D spectral numerical modelling. Furthermore no current studies have identified specific wave metrics that drive kelp morphological characteristics, which the current work will aim to achieve. Other considerations such as possible differences in morphology between sub-canopy and canopy species, and shallow and deep water kelp populations have not been considered using advanced numerical and statistical tools.

The kelps *Laminaria pallida* and *Ecklonia maxima* (Osbeck) Papenfuss are important habitat forming seaweeds that exist around the coast of South Africa and offer a unique opportunity to investigate the drivers of macroalgal morphological characteristics between canopy and sub-canopy species, as well as between deep and shallow water environments. Although both these species exist in the sub-tidal, *L. pallida* dominates deeper waters while *E. maxima* forms dense surface canopies in the sub-tidal to shallower waters. Therefore, *E. maxima* may be exposed to variations in wind driven surface waves, swell waves and temperature; while *L. pallida* is exposed to variations in swell waves and temperature. @Molloy1996-la investigated the effect of depth and wave exposure on *L. pallida* at different depths and wave exposure and showed that depth had a greater effect than wave exposure when considering all the morphological characteristics (wave exposure alters with depth); when considering individual characteristics, however, wave exposure had the most significant effect on blade thickness. Another study by @Rothman2017-pn investigated the changes in morphology in shallow populations of *L. pallida* and *E. maxima* along the South African coastline and into Namibia; *E. maxima* exhibited no morphological changes along the coast but the stipes of *L. pallida* become increasingly hollow further north along the coastline. The authors suggested that turbidity in relation to light attenuation was the environmental driver responsible for this change.

@@

Sites were chosen to represent an array of temperature and wave gradients, as indicated in figure \ref{fig:map study sites}. St. Helena Bay and Betty's Bay constituted the north western and south eastern boundary sites, respectively. These sites are roughly 300 km apart, lie within separate marine provinces, and span across the majority of the south-west coast, in varying thermal and wave energy regimes. The region is dominated by kelp communities that persist in contrasting abiotic environments. The west coast region has been termed as cool-temperate, which is defined as a region where mean monthly temperatures are always above 10°C and always below 15°C [@Smit2013a]. East of Cape Point marks the beginning of an overlap or transition area, which is also referred to as the Benguela-Agulhas Transition Zone [@Smit2013a]. The Agulhas Marine Province is characterised by a wide temperature range of up to 7°C difference between mean monthly temperatures between summer and winter and is classified as warm-temperate [@Smit2013a].

@@

The annual and seasonal patterns in both wind influence the thermal regime around the coastline by inducing upwelling. Upwelling brings deep, cool nutrient rich water to coastal areas which induces a decrease in temperature [@Field1980a; @Andrews1980; @Gill1974-nt; @Cram1970-jw; @Rouault2010-gi; @Blanke2002]. The wind speed and wind direction help drive upwelling in summer on the western side of the Cape peninsula, where southerly winds blow parallel to the coast and trigger upwelling events [@Field1980a; @Rouault2010-gi]. This is not the case for False Bay which is shielded from the dominating wind and swell the Cape Peninsula is subject to, with the exception being the winter months where upwelling becomes more favourable for the region due to the change in swell and wind direction, causing lower temperatures to dominate during this time [@Rouault2010-gi; @Andrews1980]. Due to the Cape Peninsula temperate latitude, winter months bring an increased frequency of frontal depressions that originate from the Southern Ocean [@Reason2006]. These low pressures are joined by large swells with increased wave energy. The nearshore environment, with the accompanied biota, therefore experiences high wave energy events, with increased frequency in winter [@Veitch2019]. The large peninsula acts as an obstruction for large south westerly swells, providing decreased wave energy along the west side of False Bay [@Shipley1964]. Conversely, the west coast of Cape Point is battered by these large swells. Multiple sites, therefore, exist where kelps grow in diverse temperature and wave energy climates, in close proximity. The topography and elevation along the Cape Peninsula channel and shield winds along False Bay. This is however absent in winter, where strong northerly winds are prevalent from St. Helena Bay to Betty's Bay [@Field1980a; @Jury1985-nr; @Andrews1980; @Jury1980-ov].

@@

Wave model variable data formed part of the South African Coastal Vulnerability Assessment, presented to the Department of Environmental Affairs (DEA) and produced by the Council for Scientific and Industrial Research (CSIR) [@Rautenbach2015a]. The South African wave climate was modeled via 20 spectral numerical wave models that simulated the offshore wave climate to the nearshore. The boundary conditions of these models were obtained by using the NOAA Wave Watch III (WWIII) model output, distributed via the National Centers for Environmental Prediction (NCEP) product (Office of the Director 2000; Environmental Modeling Center / Marine Modeling Branch 2005). The particular hindcast product utilised during the DEA-CSIR study spans 1994-2013 at a 3-hour resolution. These data were then used to model swell propagation into the coastal models while wind waves (seas) were generated via stationary computations in the Simulating Waves in the Nearshore SWAN model [@Booij1997-nx]. The assumption of stationary computations are acceptable as the model domains were small enough so the temporal variation of the model boundary were slower than the time it takes for that boundary condition to propagate to the coast. SWAN allows one to extract wave variables from specific gridded locations in the nearshore. For False Bay, a resolution of 200 meters was modelled and output produced at both the 7 meter and 15 meter isobaths. A 200 meter resolution was used as the False Bay computational grid was nested within a larger grid (1 kilometer resolution). This allowed for a computational effective wave resolution of increasing resolution, from the NCEP, low resolution output to nested, high resolution coastal output. For Table Bay and east of Cape Hangklip the resolution was 500 meters and also had output at the 7 meters and 15 meter isobaths. These contour outputs were chosen in the original study by the CSIR as most engineering run-up calculations require wave parameter information at these contour depths and were the main focus of the original study. For this study the 7 meter contours were used. These data were then used to calculate over all wave power (kW/m) as this was considered the best measure of overall wave exposure. Annual wave power was plotted and categorised into different wave exposure categories which ranged from fully sheltered to extremely exposed.

@@

```{r schematic, echo=FALSE, fig.cap='\\label{fig:schematic}Schematic of *L. pallida* and *E. maxima* morphology [@Dyer2018-rv].', fig.height=6, fig.width=6}

@@

Morphological measurements of *L. pallida* and *E. maxima* were collected at 18 sites along the Western Cape coast of South Africa (Fig. 1) between October 2014 and April 2015. In each instance only the thirteen largest individuals were sampled to ensure that only mature and fully grown sporophytes were measured. In instances where one of the species was not present a site, the nearest population of the missing species was sampled. Because the macroalgae differ in morphological features, species-specific morphological characteristics were included and are presented in Tables 1 and 2. Between February 2017 and November 2018, morphological measurements for *E. maxima* individuals in shallow water (<1m) at seven sites along the Western Cape coast of South Africa were also collected. The same morphological characteristic measurements were taken as for the deeper *E. maxima*, and this allowed comparison between morphological characteristics between deep and shallow individuals within sites. Measurements were not collected for *L. pallida* in the shallow depths, as this species is largely absent from the shallow in this portion of the South African coastline.

@@

Summary statistics were calculated for each site and regions for each of the environmental variables considered in this study, and are presented in Table 3. The summary statistics calculated for temperature, waves and wind variables were minimum, maximum, mean, range and standard deviation for annual timescales. In addition, annual median wave and wind direction were also calculated. The median was calculated for wind and wave direction, as issues arise when calculating the mean and standard deviation for compass metrics. Dot and whisker plots were used to present the summary statistics for both temperature and wave variables. Summary statistics for wind was not plotted and instead are discussed, as the data was course relative to the other environmental variables considered in this study.

Morphological characteristics in relation to temperature and wave exposure were investigated using non-parametric statistical methods. Morphological characteristics were categorised according to temperature regime and wave exposure category for each species and population of kelp in this study. Significant differences between temperature regimes and across wave exposure categories were tested with wilcox tests and kruskal tests respectively, and notched boxplots allowed comparison between individual categories of wave exposure. In order to investigate specific drivers of morphological characteristics a distance-based redundancy analysis was performed using the *rda* function in the *vegan* software package [@Oksanen2015-vj] in R (R Core Team, 2019). The abiotic data covered all study sites for both species of kelp. Both morphology and abiotic data (temperature, waves and wind) were standardised using the *decostand* function in the *vegan* software package; morphology data were used as response variables and the abiotic data as explanatory variables. To determine the explanatory variables that best describe patterns in the response data, a full RDA was performed using a complete set of explanatory variables. Forward selection was then used to reduce the number of explanatory variables, as well as prevent the inflation of overall type I error. To further improve the model, pairwise coefficients and Variance Inflation Factor (VIF) were calculated to identify variables with high multicollinearity. The computation of the parsimonious RDA was followed by permutation tests of the adjusted *R*^2^ to assess significance of constraints.

@@

Our data show the western side of the peninsula experiences higher significant wave heights and variation in wave heights compared to the eastern side of the peninsula on both an annual scale (see figure \ref{fig:wave variable plot}). Mean significant wave height ranged from 0.5 ± 0.2 m (mean ± SD) at St. Helena Bay to 2.3  ± 0.7 m at Olifantsbos on the western side of the peninsula, and on the eastern side of the peninsula it ranged from 0.9 ± 0.4 m at Buffelsbaai to 1.9 ± 0.6 m at Betty’s Bay (see figure \ref{fig:wave variable plot}A). The maximum significant wave height ranged from 1.4 m at St. Helena Bay to 4.7 m at Olifantsbos on the western side of the peninsula, and on the eastern side ranged from 2.6 m at Buffelsbaai to 4.2 m at Betty’s Bay. Mean peak period for sites on the western side of the peninsula ranged from 10.3 ± 3.5 s at St. Helena Bay to 11.0 ± 2.0 s at Olifantsbos, and ranged from 10.6 ± 3.0 s at Buffelsbaai to 10.8 ± 2.2 s at Betty’s Bay (see figure \ref{fig:wave variable plot}B). The data shows no trend in mean peak period for the coastline. These data show that the western side of the peninsula has a lower variation (SD) compared to the eastern side of the peninsula, with Miller’s point showing the highest variation across timescales. Maximum peak period ranged from 18.9 s at St. Helena Bay to 18.4 s at Olifantsbos for the western side of the peninsula, and ranged from 18.0 s at Buffelsbaai to 18.0 s at Betty’s Bay on the eastern side of the peninsula. When wave variables were grouped by region significant differences were found for both mean significant swell height and mean peak period. The western region The western region experiences a significantly higher mean significant swell height (p < 0.05; see figure \ref{fig:wave variable plot}C) and mean peak period (p < 0.05; see figure \ref{fig:wave variable plot}D).

@@

Data for median wave direction showed no trend across timescales; however the variation (SD) in wave direction is lower on the western side of the peninsula compared to the eastern side both annually and for winter. Summer data were the exception and showed lower variation for the eastern side of the peninsula (see figure \ref{fig:wave rose}).

@@

In Figure \ref{fig:wave power map} the total coastal wave exposure of the Cape Peninsula is given in terms of wave energy (kW per meter wave crest length). The classification from fully sheltered to extremely exposed is based on the total wave energy upper and lower limits. The western periphery of the Cape peninsula is almost continuously exposed to high wave exposures while the eastern periphery of the peninsula (western coastline of False Bay) revealed sheltered wave exposures (see figure \ref{fig:wave power map}). There the marked seasonality, with higher energy waves during winter, may be clearly observed once more.

@@

Forward selection, assessment of VIF and Pearson correlation coefficients allowed us to retained as the most parsimonious descriptors of deep *E. maxima* morphological characteristics with an adjusted *R*2 of 0.39, explaining 67% of the variation (global permutation test on final model: d.f = 5, *F* = 2.4, *p* = 0.026). RDA1 was the only significant axis in the model (p = 0.03) and explained 72% of the variation. Primary length, primary width and stipe circumference were positively influenced by *hs_mean*, *tp_mean* and *dir_median* for for kelp from the western region. The remaining morphological characteristics were positivist influenced by *min_temp* and negatively influenced by *tp_sd* for kelp from the eastern region.

>>>>

---
title: Using numerical modelling to investigate temperature and wave exposure as drivers
  of macroalgal morphological characteristics
author: "R. Coppin, C. Rautenbach, T. Ponton and A.J Smit"
bibliography: references.bib
output:
  html_document:
    df_print: paged
  pdf_document:
    fig_caption: yes
  word_document:
    reference_docx: frontiers_template.docx
geometry: margin=1in
fontsize: 10pt
papersize: A4
tables: yes
abstract: 'Macroalgal morphological variation is determined to a large extent by
  a combination of environmental factors, with wave exposure and temperature
  perhaps the main influences as they are key environmental properties to which a species becomes locally adapted.                Macroalgae have shown to exhibit different responses to different magnitudes of exposure to waves, such as reduction
  in overall size and strength increasing traits. In terms of temperature, warmer
  environments have been shown to reduce the overall size of resident and transplanted
  species. However, none of the past studies have identified specific wave and temperature
  metrics responsible for the morphological adaptation macroalgae exhibit. Past research
  has often used simple or two-dimensional models of wave exposure, which do not
  take into account important aspects of the nearshore environment such as wave breaking,
  refraction and diffraction. Furthermore, past studies have often used satellite-derived
  datasets as sources for temperature data; however, such data have been shown to have large bias when 
  applied to the nearshore environment. This study used *in situ* temperature data and wave power
  metrics calculated from a 3D-numerical model to identify specific temperature and wave metrics
  responsible for morphological adaptation of the kelp, *Ecklonia maxima* and *Laminaria pallida*. The results identify wave      exposure as the main driver of morphological adaptation while identifying specific wave metrics.
  Furthermore, the results show differences in wave metrics between species; and between
  deep and shallow populations.'
---

@@

Kelps are a group of large seaweeds of the order Laminariales (Ochrophyta), which despite their relatively low taxonomic diversity of species in genera [@Bolton2010], form the basis of one of the most productive ecosystems globally [@Mann1973].  Kelps generally have a dependence on cool-temperate and arctic seawater temperatures [@Bolton2010; @Mohring2014-bk; @Cavanaugh2011-ib; @Dayton1985], and dominate the nearshore biomass within the rocky shallow coasts in both hemispheres [@Steneck2002]. Their size and complex morphology provide a heterogeneous habitat structure [@Steneck2002] that accommodate a multitude of turf and sub-canopy seaweed species, and diverse assemblages of sessile and mobile invertebrates and vertebrates [@Mann1973; @Steneck2002], each depending on a wide suite of ecological services provided by kelp forests [@Gaines1987-bg; @Paul1993-bm; @Levin1994-qf; @Willis2003-xo; @anderson1997; @Dayton1985]. Wave exposure and temperature are regarded as important environmental drivers of kelp forests, and play a role in their distribution [@Gorman2013-nj], abundance [@Cavanaugh2011-ib; @Dayton1998-xw], diversity [@Wing2007-tv; @Wernberg2008], composition [@Leliaert2000a; @Dayton1985; @Norderhaug2012-ks; @Harley2012], growth [@Cousens1982-ck] and productivity [@Pedersen2012-vr; @Dayton1998-xw; @Krumhansl2012].

Given that kelps are cool-temperate organisms which are vulnerable to dislodgement under high wave exposure scenarios, increases in storm frequency and magnitude, and changes in ocean temperature, changes in these influences pose a direct threat to their survival; which is a concern considering that these variables are known to be affected by climate change[@Russo2014; @Reguero2019; @Weisse2010; @Rose2012; @Rummukainen2012]. Warmer temperatures reduces the resilience of kelp individuals to storm disturbance [@wernberg2010], causes fragmentation through weakening tissues [@Simonson2015] and reduces growth and productivity [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]; while storms and high wave energy dislodge and break kelps [@byrnes2011; @graham2004; @seymour1989]. Despite these threats and disturbances kelps persist across a broad range of environments which is largely due to their morphological plasticity [@FowlerWalker2006; @wernberg2010]. A study by @Wernberg2003-mp investigated the morphology of *Ecklonia radiata* (C. Agardh) J. Agardh. in order to quantify the morphological variation and whether it was dependent on spatial differences along the Australian coast. They found no correlation between spatial distance and morphological similarity; rather the morphology of the kelps was representative of multiple environmental forcings on different morphological characters at different spatial scales [@Wernberg2003-mp]. Due to the complex effects of environmental drivers on kelp morphology, one can also expect differences in morphology between deep and shallow water populations of the same species. Deep and shallow water environments may differ in abiotic processes such as the degree of water column mixing [@Smit2013a], solar heating [@Dellatorre2012; @Dunne2001] and effects of wave dampening [@Kobayashi1993-os; @Mia2013; @Dubi1995].

Although multiple environmental factors play a role in kelp morphological adaptation, wave exposure and temperature have been identified as the main drivers across various kelp species [@bekkby2014; @Wernberg2005c; @Denny2002; @Hurd2000; @FowlerWalker2005; @Thomsen2004; @mabin2013]. For example, a study by @Serisawa2002-dc compared the morphology of *Ecklonia cava* Kjellman growing at warmer and cooler sites, and showed wrinkling of the blade and reduced size to be a characteristic of higher temperature ranges. Warmer temperature affects important kelp physiological processes such as photosynthesis and respiration, which influences growth and productivity and ultimately leads to a reduction in size [@bearham2013; @gerard1997; @gao2013; @zimmerman1986]. Past research has shown that in highly wave exposed areas kelp morphology tends to take on characteristics which reduce drag, increase strength and increase flexibility [@Denny2002; @Hurd2000; @FowlerWalker2005; @Thomsen2004]. For example, a study by @Wernberg2005c examined the consistency of wave exposure as a driver of *E. radiata* across a broad geographic range, and showed trends towards drag-reducing (small size, narrow laterals and blades, low spinosity) and increased strength (large holdfast, thick stipe and thick blades and lamina) at high wave exposure sites.

Measures of wave exposure in ecological studies often incorporate integrative measures of hydrodynamic conditions at a particular site which are based on cartographical models of exposure. Cartographical models of wave exposure are 'fetch-based-models' which measure the length of open water associated with a particular site in a straight line, and are regarded as simple measures of wave exposure. Advances in numerical modelling based on physical/linear wave theory incorporate more complexity (wind forcing, wave-wave interactions, wave breaking, diffraction and variation in wave direction) into the models and allow for a quantitative, reproducible approach for characterising the hydrodynamic environment. Although @mabin2013 investigated kelp morphological characteristics using a 2D WAM model, currently no studies investigating macroalgal morphological characteristics in relation to the wave environment have incorporated 3D spectral numerical modelling. Furthermore no studies have identified specific wave metrics that drive kelp morphological characteristics, which the current work will aim to achieve. Other considerations such as possible differences in morphology between sub-canopy and canopy species, and shallow and deep water kelp populations have not been considered using advanced numerical and statistical tools.

The kelps *Laminaria pallida* Greville and *Ecklonia maxima* (Osbeck) Papenfuss are important habitat forming seaweeds that exist around the coast of South Africa and offer a unique opportunity to investigate the drivers of macroalgal morphological characteristics between canopy and sub-canopy species, as well as between deep and shallow water environments. Although both these species exist in the sub-tidal, *L. pallida* dominates deeper waters while *E. maxima* forms dense surface canopies in the shallow sub-littoral. Therefore, *E. maxima* may be exposed to variations in wind driven surface waves, swell waves and temperature, while *L. pallida* is exposed to variations in swell waves and temperature. @Molloy1996-la investigated the effect of depth and wave exposure on *L. pallida* at different depths and wave exposure and showed that depth had a greater effect than wave exposure when considering all the morphological characteristics; when considering individual characteristics, however, wave exposure had the most significant effect on blade thickness. Another study by @Rothman2017-pn investigated the changes in morphology in shallow populations of *L. pallida* and *E. maxima* along the South African coastline and into Namibia; *E. maxima* exhibited no morphological changes along the coast but the stipes of *L. pallida* become increasingly hollow further north along the coastline. The authors suggested that turbidity in relation to light attenuation was the environmental driver responsible for this change.

@@

Sites were chosen to represent an array of temperature and wave gradients, as indicated in figure \ref{fig:map study sites}. St. Helena Bay and Betty's Bay constituted the north-western and south-eastern boundary sites, respectively. These sites are roughly 300 km apart, lie within separate marine provinces @Smit2017, and span the majority of the south-west coast, in varying thermal and wave energy regimes. The region is dominated by kelp communities that persist in contrasting abiotic environments. The west coast region has been termed cool-temperate, which is defined as a region where mean monthly temperatures are always above 10°C and below 15°C [@Smit2013a]. East of Cape Point marks the beginning of an overlap or transition area, which is also referred to as the Benguela-Agulhas Transition Zone [@Smit2013a]. The Agulhas Marine Province is characterised by a wide temperature range of up to 7°C difference between mean monthly temperatures between summer and winter and is classified as warm-temperate [@Smit2013a].

@@

The annual and seasonal patterns in both wind influence the thermal regime around the coastline by inducing upwelling. Upwelling brings deep, cool nutrient-rich water to coastal areas which causes a decrease in temperature [@Field1980a; @Andrews1980; @Gill1974-nt; @Cram1970-jw; @Rouault2010-gi; @Blanke2002]. The wind speed and wind direction help drive upwelling in summer on the western side of the Cape Peninsula, where southerly winds blow parallel to the coast and trigger upwelling events [@Field1980a; @Rouault2010-gi]. This is not the case for False Bay, which is shielded from the dominating wind and swell that the Cape Peninsula is subjected to. In the winter months, upwelling becomes more favourable in False Bay due to the change in swell and wind direction, causing lower temperatures to dominate during this time [@Rouault2010-gi; @Andrews1980]. Due to the Cape Peninsula temperate latitude, winter months bring an increased frequency of frontal depressions that originate from the Southern Ocean [@Reason2006]. These low pressures are joined by large swells with increased wave energy. The nearshore environment, with the accompanied biota, therefore experiences high wave energy events, with increased frequency in winter [@Veitch2019]. The large peninsula acts as an obstruction for large south westerly swells, providing decreased wave energy along the west side of False Bay [@Shipley1964]. Conversely, the west coast of Cape Point is battered by these large swells. Multiple sites, therefore, exist where kelps grow in diverse temperature and wave energy climates, in close proximity and on the same side. The topography and elevation along the Cape Peninsula channel and shield winds along False Bay. This is, however, absent in winter, when strong northerly winds are prevalent from St. Helena Bay to Betty's Bay [@Field1980a; @Jury1985-nr; @Andrews1980; @Jury1980-ov].

@@

Wave model variable data formed part of the South African Coastal Vulnerability Assessment, presented to the Department of Environmental Affairs (DEA) and produced by the Council for Scientific and Industrial Research (CSIR) [@Rautenbach2015a]. The South African wave climate was modeled via 20 spectral numerical wave models that simulated the offshore wave climate to the nearshore. The boundary conditions of these models were obtained by using the NOAA Wave Watch III (WWIII) model output, distributed via the National Centers for Environmental Prediction (NCEP) product (Office of the Director 2000; Environmental Modeling Center / Marine Modeling Branch 2005). The particular hindcast product utilised during the DEA-CSIR study spans 1994-2013 at a 3-hour resolution. These data were then used to model swell propagation into the coastal models while wind waves (seas) were generated via stationary computations in the Simulating Waves in the Nearshore SWAN model [@Booij1997-nx]. The assumption of stationary computations is acceptable as the model domains were small enough so the temporal variation of the model boundary was slower than the time it takes for that boundary condition to propagate to the coast. SWAN allows one to extract wave variables from specific gridded locations in the nearshore. For False Bay, a resolution of 200 meters was modelled and output produced at both the 7 meter and 15 meter isobaths. A 200 meter resolution was used as the False Bay computational grid was nested within a larger 1 kilometer resolution grid. This allowed for a computational effective wave resolution of increasing resolution, from the NCEP, low resolution output to nested, high resolution coastal output. For Table Bay and east of Cape Hangklip the resolution was 500 meters and also had output at the 7 meters and 15 meter isobaths. These contour outputs were chosen in the original study by the CSIR as most engineering run-up calculations require wave parameter information at these contour depths and were the main focus of the original study. For this study the 7 meter contours were used. These data were then used to calculate over all wave power (kW/m) as this was considered the best measure of overall wave exposure. Annual wave power was plotted and categorised into different wave exposure categories which ranged from fully sheltered to extremely exposed.

@@

```{r schematic, echo=FALSE, fig.cap='\\label{fig:schematic}Schematic of *Laminaria pallida* and *Ecklonia maxima* morphology [@Dyer2018-rv].', fig.height=6, fig.width=6}

@@

Morphological measurements of *L. pallida* and *E. maxima* were collected at 18 sites along the Western Cape coast of South Africa (Fig. 1) between October 2014 and April 2015. In each instance only the 13 largest individuals were sampled to ensure that only mature and fully grown sporophytes were measured. In instances where one of the species was not present a site, the nearest population of the missing species was sampled. Because the macroalgae differ in morphological features, species-specific morphological characteristics were included and are presented in Tables 1 and 2. Between February 2017 and November 2018, morphological measurements for *E. maxima* individuals in shallow water (<1m) at seven sites along the Western Cape coast of South Africa were also collected. The same morphological characteristic measurements were taken as for the deeper *E. maxima*, and this allowed comparison between morphological characteristics between deep and shallow individuals within sites. Measurements were not collected for *L. pallida* in the shallow depths, as this species is largely absent from the shallow in this portion of the South African coastline.

@@

Summary statistics were calculated for each site and region for each of the environmental variables considered in this study, and are presented in Table 3. In addition, annual median wave and wind direction were also calculated. The median was calculated for wind and wave direction, as issues arise when calculating the mean and standard deviation for compass metrics. Summary statistics for wind was not plotted and instead are discussed, as the data were course relative to the other environmental variables considered in this study.

Morphological characteristics in relation to temperature and wave exposure were investigated using non-parametric statistical methods. Morphological characteristics were categorised according to temperature regime and wave exposure category for each species and population of kelp in this study. Significant differences between temperature regimes and across wave exposure categories were tested with Wilcox tests and Kruskal tests, respectively, and notched boxplots allowed comparison between individual categories of wave exposure. In order to investigate specific drivers of morphological characteristics a distance-based redundancy analysis (db-RDA) was performed using the *rda* function in the *vegan* software package [@Oksanen2015-vj] in R (R Core Team, 2019). The abiotic data covered all study sites for both species of kelp. Both morphology and abiotic data (temperature, waves and wind) were standardised using the *decostand* function in the *vegan* software package; morphology data were used as response variables and the abiotic data as explanatory variables. To determine the explanatory variables that best describe patterns in the response data, a full RDA was performed using a complete set of explanatory variables. Forward selection was then used to reduce the number of explanatory variables, as well as prevent the inflation of overall Type I error. To further improve the model, pairwise coefficients and Variance Inflation Factor (VIF) were calculated to identify variables with high multicollinearity. The computation of the parsimonious RDA was followed by permutation tests of the adjusted *R*^2^ to assess significance of constraints.

@@

Our data show the western side of the peninsula experiences higher significant wave heights and variation in wave heights compared to the eastern side of the peninsula on both an annual scale (see figure \ref{fig:wave variable plot}). Mean significant wave height ranged from 0.5 ± 0.2 m (mean ± SD) at St. Helena Bay to 2.3  ± 0.7 m at Olifantsbos on the western side of the peninsula, and on the eastern side of the peninsula it ranged from 0.9 ± 0.4 m at Buffelsbaai to 1.9 ± 0.6 m at Betty’s Bay (see figure \ref{fig:wave variable plot}A). The maximum significant wave height ranged from 1.4 m at St. Helena Bay to 4.7 m at Olifantsbos on the western side of the peninsula, and on the eastern side ranged from 2.6 m at Buffelsbaai to 4.2 m at Betty’s Bay. Mean peak period for sites on the western side of the peninsula ranged from 10.3 ± 3.5 s at St. Helena Bay to 11.0 ± 2.0 s at Olifantsbos, and ranged from 10.6 ± 3.0 s at Buffelsbaai to 10.8 ± 2.2 s at Betty’s Bay (see figure \ref{fig:wave variable plot}B). The data show no trend in mean peak period for the coastline. These data show that the western side of the peninsula has a lower variation (SD) compared to the eastern side, with Miller’s Point showing the highest variation across timescales. Maximum peak period ranged from 18.9 s at St. Helena Bay to 18.4 s at Olifantsbos for the western side of the peninsula, and ranged from 18.0 s at Buffelsbaai to 18.0 s at Betty’s Bay on the eastern side of the peninsula. When wave variables were grouped by region significant differences were found for both mean significant swell height and mean peak period. The western region experiences a significantly higher mean significant swell height (p < 0.05; see figure \ref{fig:wave variable plot}C) and mean peak period (*p* < 0.05; see figure \ref{fig:wave variable plot}D).

@@

Data for median wave direction showed no trend across timescales; however, the variation (SD) in wave direction is lower on the western side of the peninsula compared to the eastern side both annually and for winter. Summer data were the exception and showed lower variation for the eastern side of the peninsula (see figure \ref{fig:wave rose}).

@@

In Figure \ref{fig:wave power map} the total coastal wave exposure of the Cape Peninsula is given in terms of wave energy (kW per meter wave crest length). The classification from fully sheltered to extremely exposed is based on the total wave energy upper and lower limits. The western periphery of the Cape Peninsula is almost continuously exposed to high wave exposures while the eastern periphery of the peninsula (western coastline of False Bay) revealed sheltered wave exposures (see figure \ref{fig:wave power map}). There the marked seasonality, with higher energy waves during winter, may be clearly observed once more.

@@

Forward selection, assessment of VIF and Pearson correlation coefficients allowed us to retained as the most parsimonious descriptors of deep *E. maxima* morphological characteristics with an adjusted *R*^2^ of 0.39, explaining 67% of the variation (global permutation test on final model: d.f = 5, *F* = 2.4, *p* = 0.026). RDA1 was the only significant axis in the model (*p* = 0.03) and explained 72% of the variation. Primary length, primary width and stipe circumference were positively influenced by *hs_mean*, *tp_mean* and *dir_median* for for kelp from the western region. The remaining morphological characteristics were positivist influenced by *min_temp* and negatively influenced by *tp_sd* for kelp from the eastern region.
